Critical assessment of Linear-Response Excitations in Molecular system\\ or \\ On the Localization of the Linear-Response excitations in Open Systems

### 2018-01-23 LRDraft.tex
update

%% Useful packages
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{mathtools,braket}
\usepackage{xifthen}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}

@@

\title{Critical assessment of Linear-Response Excitations in Molecular system\\
or \\
On the Localization of the Linear-Response excitations in Open Systems}
\author{You}

@@

We start this analysis by deriving an ulterior expression of the response density operator in the frequency space. We consider equation \eqref{rhoPrimeOmegaDef1} and observe that, for any 
operator $\op O$, we may write:  
\be
\commutator{\dmnot}{\op O} = \dmnot \op O \op Q_0 - \op Q_0 \op O \dmnot\;,
\ee
where $\op Q_0= \identity - \dmnot$ is the projection operator in the subspace orthogonal to $\dmnot$. So, thanks to this relation we have that:
\be
\dm'(\omega) = \ii\sum_p\left(\ketbra{\psi_p}{\psi_p}\op A(\omega)\op Q_0 - \op Q_0\op A(\omega)\ketbra{\psi_p}{\psi_p}\right) \;.
\ee
This equation suggests us to define the \emph{fluctuation states}:
\be\lb{fluctuationStateDef1}
\ket{f_p(\omega)} \doteq -\ii\op Q_0\op A(\omega)\ket{\psi_p} \;, \qq
\bra{f_p(\omega)} \doteq \bra{\psi_p}\ii\op A(-\omega)\op Q_0 \;,
\ee
where the $\omega$ dependence of the bra is consistent with hermitian conjugation property of the Fourier transform of a hermitian operator. An inspection of equation \eqref{ketBraTimeEvol1} 
evidence that formulas \eqref{fluctuationStateDef1} are equivalent to the action of the $\op Q_0$ projector acting on the time evolution of the (occupied) eigenstates of $\hnot$, written 
in the $\omega$ domain. Fluctuation states are, by definition, orthogonal to the occupied eigenstates of $\hnot$, so that:
\be
\brket{f_p(\omega)}{\psi_q} = \brket{\psi_q}{f_p(\omega)} = 0 \;, \forall \, p,q,\omega \;.
\ee
The response density written in terms of the fluctuation states reads:
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p(-\omega)} + \ketbra{f_p(\omega)}{\psi_p}\right) \;,
\ee
and consequently the linear response functional can be expressed as:
\be\lb{LinearResponseFunctDef2}
{\cal R}_{\op O}(\omega) = 
\sum_p\left(\bra{f_p(-\omega)}\op O\ket{\psi_p} + \bra{\psi_p}\op O\ket{f_p(\omega)}\right) \;.
\ee
The task of the linear response formalism to provided reliable results for the expectation value of $\op O$ can be reformulated in the problem of finding an accurate and unbiased
expression for the fluctuation states in the domain in which $\op O\ket{\psi_p}$ (and its hermitian conjugate) are not vanishing.

\subsection{Localization argument for the fluctuation states}
\lb{LocalizationFluctuationStates}

Starting from the Liouville equation \eqref{LiouvillianRhopomegaDef3} it is possible to derive a set of equation of motion for the fluctuation states. 
The following equations are useful for this task: 
\begin{align}
\Liouvnot\dm'(\omega) & = \sum_p\left(-\ketbra{\psi_p}{f_p(-\omega)}\left(\hnot-\eps_p\right) + \left(\hnot-\eps_p\right)\ketbra{f_p(\omega)}{\psi_p}\right) \;, \nn \\
\coupl\dm'(\omega) &= \commutator{\op V'(\omega)}{\dmnot} = \op Q_0 \op V'(\omega) \dmnot - \dmnot \op V'(\omega) \op Q_0 = \nn \\
& = \sum_p\left( \op Q_0 \op V'(\omega) \ketbra{\psi_p}{\psi_p} - \ketbra{\psi_p}{\psi_p} \op V'(\omega) \op Q_0 \right) \;, \nn \\
\commutator{\op \Phi(\omega)}{\dmnot} & =  \sum_p\left( \op Q_0 \op \Phi(\omega) \ketbra{\psi_p}{\psi_p} - \ketbra{\psi_p}{\psi_p} \op \Phi(\omega) \op Q_0 \right) \;. \nn
\end{align} 
Liouville equation written in this fashion implies the following equations for the fluctuation states, that can be extracted by multiplying from left and right for the 
occupied eigenstates of $\hnot$:
\begin{align}\lb{fluctuationStateEqMotion1}
\left[\bbomega - \left(\hnot-\eps_p\right)\right]\ket{f_p(\omega)} &= \op Q_0\left(\op V'(\omega) + \op \Phi(\omega)\right)\ket{\psi_p} \;, \nn \\
\bra{f_p(-\omega)}\left[-\bbomega - \left(\hnot-\eps_p\right)\right] &= \bra{\psi_p}\left(\op V'(\omega) + \op \Phi(\omega)\right)\op Q_0 \;.
\end{align}
These equations allows us to present a localization argument for the fluctuation states. We discuss this point for the right vectors, \emph{i.e.} for the first line of 
\eqref{fluctuationStateEqMotion1}, but the same argument holds also for the left ones.

We split the unperturbed Hamiltonian as the kinetic $\op T$ plus the potential $\op V$ term, so that the equation of motion for $\ket{f_p(\omega)}$ is recasted as:
\be\lb{fluctuationStateEqMotion}
\left(\bbomega+\eps_p -\op T\right)\ket{f_p(\omega)} = \op V\ket{f_p(\omega)} + \op Q_0\left(\op V'(\omega) + \op \Phi(\omega)\right)\ket{\psi_p} \;.
\ee
This equation has the formal structure of an inhomogeneous Helmholtz equation and can be written in the coordinate representation as:
\be
\left(k^2_p(\omega)+\ii\eta + \nabla^2\right)f_p(\omega,\r) = s(\omega,\r) \;\;, {\rm with} \;\; k^2_p(\omega) = 2(\omega+\eps_p) \;,
\ee
and admits an integral representation in the form
\be\lb{fluctuationStateIntEq1}
f_p(\omega,\r) = \int \dd\r' g_H(\r-\r')s(\r') \;,
\ee
where $g_H(\r)$ is the Green function of the Helmholtz operator in three dimensions. This integral representation allows us to assess the localization character 
of the fluctuation states, \emph{i.e.} the behavior of $f_p(\omega,\r)$ when $\r$ becomes large. Indeed, assuming that the support of the source term is compact
we have that, for $\r$ outside the domain in which the integrand of \eqref{fluctuationStateIntEq1} is not vanishing:
\be\lb{fluctuationStateIntEq2}
f_p(\omega,\r) \sim \frac{e^{\ii \sqrt{k^2_p(\omega)} |\r|}}{|\r|} \;,
\ee
so the asymptotic character of the fluctuation states depends on the sign of $k^2_p(\omega)$. In particular equation \eqref{fluctuationStateIntEq2} describes an
exponential damping for $\omega < |\eps_p|$ and an oscillating power law behavior for $\omega \geq |\eps_p|$.

This argument holds for all the fluctuation states that contribute to the perturbation-induced expectation value \eqref{LinearResponseFunctDef2}. We conclude that
the linear response functional is built in terms of pure localized functions for $\omega < |\eps_h|$, instead in the range $\eps_h|\leq\omega<|\eps_0|$ it contains both
contributions from local and non-local quantities and, finally is composed only by non-local fluctuation states for $\omega\geq|\eps_0|$.

\subsection{Criterion for the choice of the underlying computational basis}

Let us further investigate formula \eqref{LinearResponseFunctDef2} pointing out some computation aspects. We introduce a computational basis $\{\ket{\mu}\}$, composed by a \emph{finite}
set of functions on which we expand the formal scalar products appearing in \eqref{LinearResponseFunctDef2} as follows:
\be\lb{LinearResponseFunctDef3}
\bra{f_p(-\omega)}\op O\ket{\psi_p} = \sum_\mu \brket{f_p(-\omega)}{\mu}\bra{\mu}\op O\ket{\psi_p} \;.
\ee
The quality of the basis can be evaluated by the requirement that the numerical coefficients appearing in the r.h.s of \eqref{LinearResponseFunctDef3} provide a reliable reconstruction 
of the formal scalar product. Here we are assuming that the ground state properties of the unperturbed system have been computed in the same basis introduced above. This means that the 
feature of the basis, that for instance for real space basis are given by the grid spacing and number of points (that determines the dimension simulation box), are chosen so that the 
coefficients $\brket{\mu}{\psi_p}$ represent a valid discretization of the occupied orbitals. The same can be stated also for the matrix elements $\bra{\mu}\op O\ket{\psi_p}$, by assuming 
that $\op O$ acts as a local operator on the occupied orbitals.

The most interesting point of this discussion concerns with the assessment of the reliable sampling of the fluctuation states through the coefficients $\brket{f_p(-\omega)}{\mu}$ and
$\brket{\mu}{f_p(\omega)}$. The question is non-trivial, as the behavior of the fluctuation states strongly depends on $\omega$, as previously explained.
The first step needed to address this problem concerns with the evaluation of the reliability of the $\{\ket{\mu}\}$ basis for $\omega=0$, \emph{i.e.} for a perturbing field that does 
not depend on time. In this case the fluctuation states, as well as the expectation value of a given observable, can be computed performing a ground state calculation in which the perturbing 
field $\op\Phi$ is added to the external potential of $\hnot$\footnote{We are interested in the linear response, so the magnitude of $\op\Phi$ has to be chosen in a way that respects this 
regime. To check this point we can compute the perturbation-induced modification of an observable an verify that its value scales linearly with the magnitude of the field}. 
The goodness of the $\{\ket{\mu}\}$ basis can thus be verified as in the standard ground state calculation, for instance by checking that, given a certain level of completeness of the basis,
the energy of the occupied (perturbed) orbitals is stable with respect to further refinements.

Once that a basis with acceptable features has been chosen and verified for $\omega=0$ we can reasonably extend its reliability to $\omega>0$ by using the localization argument presented above. 
Indeed, we have argued that fluctuation states retain a local behavior up to $\omega=|\eps_h|$ and so we can expect that, up to this threshold value, the $\{\ket{\mu}\}$ basis represents an 
effective tool for a precise discretization of \eqref{LinearResponseFunctDef3}, thus providing a reliable estimate of the linear response.

@@

\subsection{Excitation operator and the analytic structure of the linear response}

@@

\subsubsection{Definition of the basis of transitions}

We now introduce the notion of \emph{basis of transitions} and identify the operators than can be expanded in such a basis.

We split the complete
set of eigenstates of $\hnot$ in two sets. The first one is composed by the eigenvectors of $\hnot$ in the subspace of the occupied orbitals. The second 
one is composed by the remaining eigenvectors, both in the bound and in the continuum region, and its elements are denoted as $\ket{\psi(\alpha)}$, where $\alpha$ 
is a multi-index in the sense described in \ref{BasicDef}. The elements of these sets satisfy the following orthonormality relation: 
\be\lb{orthonormalityDef1}
\brket{\psi(\alpha)}{\psi_p} = \brket{\psi_p}{\psi(\alpha)} = 0 \;, \qq \brket{\psi_p}{\psi_q} = \delta_{pq} \;, \qq
\brket{\psi(\alpha)}{\psi(\beta)} = \delta(\alpha-\beta) \;.
\ee
Obviously, the direct sum of these two sets coincides with the complete set of eigenstates of $\hnot$ and thus 
constitute a complete basis. 
A generic operator can be decomposed in this basis and its formal expansion contains four distinct contribution of the form $\ketbra{\psi_p}{\psi_q}$, 
$\ketbra{\psi(\alpha)}{\psi(\beta)}$, $\ketbra{\psi(\alpha)}{\psi_p}$, $\ketbra{\psi_p}{\psi(\alpha)}$. However an inspection of equations \eqref{rhoPrimeOmegaDef1}
and \eqref{LiouvillianRhopomegaDef3} evidences that the relevant operators in the contest of the linear response are expressed as a commutator with $\dmnot$. 
For this class of operators the only non vanishing contribution comes from the \emph{transition operators}:
\be
\excite{p}{\alpha} \equiv \ketbra{\psi(\alpha)}{\psi_p}\;, \qq
\decay{\alpha}{p} \equiv \ketbra{\psi_p}{\psi(\alpha)} \;.
\ee
The orthonormality conditions \eqref{orthonormalityDef1} imply that:
\be\lb{transitionOpOrthonormalityDef1}
\trace{\excite{p}{\alpha}\excite{q}{\beta}} = \trace{\decay{\alpha}{p}\decay{\beta}{q}} = 0 \;, \qq
\trace{\excite{p}{\alpha}\decay{\beta}{q}} = \delta_{pq}\delta(\alpha-\beta) \;,
\ee
so that the expansion of a suitable operator $\op O$ in the basis of transitions reads:
\be\lb{OperatorTransBasisExpDef1}
\op O = \sum_p\sint \dd \alpha\left(
\excite{p}{\alpha}\trace{\decay{\alpha}{p}\op O} + \decay{\alpha}{p}\trace{\excite{p}{\alpha}\op O}
\right) \;.
\ee
In order to recast this expansion in a simpler expression it turns out to be useful to define the rank-2 row vectors of the transition basis: 
\be
\sbra{p\alpha}{\cdot} \equiv 
\mat{ \excite{p}{\alpha} \;, & \decay{\alpha}{p} }\;,
\ee
and we also introduce a vectorial notation to deal with the coefficients of the expansion of an operator in the transition basis, that is: 
\be\lb{OperatorTransitionCoeffDef1}
\sket{p\alpha}{\op O}  \equiv \mat{ \trace{\decay{\alpha}{p}\op O } \\ 
\trace{\excite{p}{\alpha}\op O}} \; .
\ee
Thanks to this formalism we have that equation \eqref{OperatorTransBasisExpDef1} can be expressed as:
\be\lb{OperatorTransBasisExpDef2}
\op O = \sum_p\sint \dd \alpha \sbra{p\alpha}{\cdot}\sket{p\alpha}{\op O} = 
\dbraket{\cdot}{\op O} \;,
\ee
where the last equality implicitly define the ulterior compact notation of the ``big-shaped'' vectors in which the summed indexes are omitted.  
In the same fashion it will be also useful for the subsequent analysis to introduce a ``conjugate'' notation, in which the basis operators are expressed
as column vector and the coefficients of the expanded operator as row ones, that is: 
\be\lb{TransitionBasisConjugateDef1}
\sket{p\alpha}{\cdot} \equiv 
\mat{\decay{\alpha}{p} \\  \excite{p}{\alpha}} \;, \qq
\sbra{p\alpha}{\op O}  \equiv 
\mat{ \trace{\excite{\alpha}{p}\op O } \;, & \trace{\decay{p}{\alpha}\op O} }
\;,
\ee
where the flip in the role of $\op e$ and $\op d$ in \eqref{TransitionBasisConjugateDef1} follows from the properties of excitation and decay operators
under hermitian-conjugate operation. The expansion of a suitable operator in the transition basis using the conjugate description reads: 
\be\lb{OperatorTransBasisConjugateExpDef1}
\op O = \sum_p\sint \dd \alpha \sbra{p\alpha}{\op O}\sket{p\alpha}{\cdot} = 
\dbraket{\op O}{\cdot} \;.
\ee

>>>>

%% Useful packages
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{mathtools,braket}
\usepackage{xifthen}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{authblk}

@@

\title{Critical assessment of Linear-Response Excitations in Molecular system\\
or \\
On the Localization of the Linear-Response excitations in Open Systems}
\author[1]{Luigi Genovese}
\author[2]{Marco D'Alessandro}
\affil[1]{Laboratoire de Simulation Atomistique (LSim), SP2M, INAC, CEA-UJF, 17 Av. des Martyrs,
38054 Grenoble, France}
\affil[2]{Istituto di Struttura della Materia-CNR (ISM-CNR), Via del Fosso del Cavaliere 100, 00133 Roma, Italia}

@@

We start this analysis by deriving an ulterior expression of the response density operator in the frequency space. We consider equation \eqref{rhoPrimeOmegaDef1} and observe that, for any 
operator $\op O$, we may write:  
\be
\commutator{\dmnot}{\op O} = \dmnot \op O \op Q_0 - \op Q_0 \op O \dmnot\;,
\ee
where $\op Q_0= \identity - \dmnot$ is the projection operator in the subspace orthogonal to $\dmnot$. So, thanks to this relation we have that:
\be
\dm'(\omega) = \ii\sum_p\left(\ketbra{\psi_p}{\psi_p}\op A(\omega)\op Q_0 - \op Q_0\op A(\omega)\ketbra{\psi_p}{\psi_p}\right) \;.
\ee
This equation suggests us to define the \emph{fluctuation states}:
\be\lb{fluctuationStateDef1}
\ket{f_p(\omega)} \doteq -\ii\op Q_0\op A(\omega)\ket{\psi_p} \;, \qq
\bra{f_p(\omega)} \doteq \bra{\psi_p}\ii\op A(-\omega)\op Q_0 \;,
\ee
where the $\omega$ dependence of the bra is consistent with hermitian conjugation property of the Fourier transform of a hermitian operator. An inspection of equation \eqref{ketBraTimeEvol1} 
evidence that formulas \eqref{fluctuationStateDef1} are equivalent to the action of the $\op Q_0$ projector acting on the time evolution of the (occupied) eigenstates of $\hnot$, written 
in the $\omega$ domain. Fluctuation states are, by definition, orthogonal to the occupied eigenstates of $\hnot$, so that:
\be
\brket{f_p(\omega)}{\psi_q} = \brket{\psi_q}{f_p(\omega)} = 0 \;, \forall \, p,q,\omega \;.
\ee
The response density written in terms of the fluctuation states reads:
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p(-\omega)} + \ketbra{f_p(\omega)}{\psi_p}\right) \;,
\ee
and consequently the linear response functional can be expressed as:
\be\lb{LinearResponseFunctDef2}
{\cal R}_{\op O}(\omega) = 
\sum_p\left(\bra{f_p(-\omega)}\op O\ket{\psi_p} + \bra{\psi_p}\op O\ket{f_p(\omega)}\right) \;.
\ee
Let us further investigate this formula pointing out some of the  computational aspects needed to explicitly perform the scalar products appearing in 
\eqref{LinearResponseFunctDef2}. 
We introduce a computational $\mu$-basis, composed by a finite set of $n_\mu$ basis elements $\{\ket{\mu}\}$, and assume that a reliable solution of the ground state problem has 
been provided in this basis. This means that the feature of the basis (that, for instance, for real space basis are given by the grid spacing and number of points) are chosen so 
that the coefficients $\brket{\mu}{\psi_p}$ represent a valid discretization of the occupied orbitals. The same can be stated also for the matrix elements $\bra{\mu}\op O\ket{\psi_p}$, 
under the hypothesis that the observable $\op O$ acts as a local operator on the occupied orbitals.

The formal scalar products that express the linear response functional \eqref{LinearResponseFunctDef1} can be evaluated in the same basis introduced above, that is: 
%For instance the first addend of \eqref{LinearResponseFunctDef2} reads: 
%\be\lb{LinearResponseFunctDef3}
%\bra{f_p(-\omega)}\op O\ket{\psi_p} = \sum_\mu \brket{f_p(-\omega)}{\mu}\bra{\mu}\op O\ket{\psi_p} \;.
%\ee
\be\lb{LinearResponseFunctDef3}
{\cal R}_{\op O}(\omega) = 
\sum_p\sum_\mu\left(\brket{f_p(-\omega)}{\mu}\bra{\mu}\op O\ket{\psi_p} + \bra{\psi_p}\op O\ket{\mu}\brket{\mu}{f_p(\omega)}\right) \;.
\ee
The task of the linear response formalism to provide reliable results for the expectation value of $\op O$ can be reformulated in the problem of finding an accurate and unbiased
expression for the numerical coefficients appearing in \eqref{LinearResponseFunctDef3}. The features of the computational basis described above ensures that this requirement should
be satisfied for $\bra{\mu}\op O\ket{\psi_p}$ (and its hermitian conjugates). 
On the contrary, the problem of the reliable reconstruction of the fluctuation states through the sampling of the coefficients $\brket{f_p(-\omega)}{\mu}$ and $\brket{\mu}{f_p(\omega)}$
turns out to be highly non-trivial, since the behavior of the fluctuation states strongly depends on $\omega$, as will be readily clarified in the next section.

%On the contrary, the most interesting point of this discussion concerns with the assessment of the 
%reliable sampling of the fluctuation states through the coefficients $\brket{f_p(-\omega)}{\mu}$ and $\brket{\mu}{f_p(\omega)}$. The problem turns out to be highly is non-trivial 
%since the behavior of the fluctuation states strongly depends on $\omega$, as will be readily clarified in the next section.

%$\brket{f_p(-\omega)}{\mu}$ and $\brket{\mu}{f_p(\omega)}$ in the domain in which the matrix elements $\bra{\mu}\op O\ket{\psi_p}$ (and its hermitian 
%conjugates) are not vanishing. 
%The task of the linear response formalism to provide reliable results for the expectation value of $\op O$ can be reformulated in the problem of finding an accurate and unbiased
%expression for the coefficients $\brket{f_p(-\omega)}{\mu}$ and $\brket{\mu}{f_p(\omega)}$ in the domain in which the matrix elements $\bra{\mu}\op O\ket{\psi_p}$ (and its hermitian 
%conjugates) are not vanishing.

% We assume that the ground state problem has been already solved and that the occupied orbitals  introduce a computational basis $\{\ket{\mu}\}$, composed by a \emph{finite}
% set of functions
% 
% Here we are dealing with a linear response problem and

% The quality of the basis can be evaluated by the requirement that the numerical coefficients appearing in the r.h.s of \eqref{LinearResponseFunctDef3} provide a reliable reconstruction 
% of the formal scalar product.

% Here we are assuming that the ground state properties of the unperturbed system have been computed in the same basis introduced above. This means that the 
% feature of the basis, that for instance for real space basis are given by the grid spacing and number of points (that determines the dimension simulation box), are chosen so that the 
% coefficients $\brket{\mu}{\psi_p}$ represent a valid discretization of the occupied orbitals. The same can be stated also for the matrix elements $\bra{\mu}\op O\ket{\psi_p}$, by assuming 
% that $\op O$ acts as a local operator on the occupied orbitals.

% aggiungere un preambolo a quello che segue. Soluzione usando le equazioni di Sternheimer e approccio con le eccitazioni....

\subsection{Sternheimer equations and localization properties of the fluctuation states}
%\subsection{Localization argument for the fluctuation states}
\lb{LocalizationFluctuationStates}

Starting from the Liouville equation \eqref{LiouvillianRhopomegaDef3} it is possible to derive a set of equation of motion for the fluctuation states. 
The following equations are useful for this task: 
\begin{align}
\Liouvnot\dm'(\omega) & = \sum_p\left(-\ketbra{\psi_p}{f_p(-\omega)}\left(\hnot-\eps_p\right) + \left(\hnot-\eps_p\right)\ketbra{f_p(\omega)}{\psi_p}\right) \;, \nn \\
\coupl\dm'(\omega) &= \commutator{\op V'(\omega)}{\dmnot} = \op Q_0 \op V'(\omega) \dmnot - \dmnot \op V'(\omega) \op Q_0 = \nn \\
& = \sum_p\left( \op Q_0 \op V'(\omega) \ketbra{\psi_p}{\psi_p} - \ketbra{\psi_p}{\psi_p} \op V'(\omega) \op Q_0 \right) \;, \nn \\
\commutator{\op \Phi(\omega)}{\dmnot} & =  \sum_p\left( \op Q_0 \op \Phi(\omega) \ketbra{\psi_p}{\psi_p} - \ketbra{\psi_p}{\psi_p} \op \Phi(\omega) \op Q_0 \right) \;. \nn
\end{align} 
Liouville equation written in this fashion implies the following equations for the fluctuation states, that can be extracted by multiplying from left and right for the 
occupied eigenstates of $\hnot$:
\begin{align}\lb{fluctuationStateEqMotion1}
\left[\bbomega - \left(\hnot-\eps_p\right)\right]\ket{f_p(\omega)} &= \op Q_0\left(\op V'(\omega) + \op \Phi(\omega)\right)\ket{\psi_p} \;, \nn \\
\bra{f_p(-\omega)}\left[-\bbomega - \left(\hnot-\eps_p\right)\right] &= \bra{\psi_p}\left(\op V'(\omega) + \op \Phi(\omega)\right)\op Q_0 \;.
\end{align}
Equations \eqref{fluctuationStateEqMotion1} have the form of a Sternheimer equation. Once projected in the $\{\ket{\mu}\}$ basis, and fixed the value of $\omega$, they can be solved 
using a self-consistent procedure (add references). This approach provides a direct evaluation of the fluctuation states and thus the linear response functional \eqref{LinearResponseFunctDef3}
can be computed.

Much more interestingly for the main topic of the present discussion, and regardless from the detailed strategy of their solutions, equations \eqref{fluctuationStateEqMotion1} 
allow us to present a \emph{localization argument} for the fluctuation states. We discuss this point for the right vectors, \emph{i.e.} for the first line of \eqref{fluctuationStateEqMotion1}, 
but the same argument holds also for the left ones. We split the unperturbed Hamiltonian as the sum of the kinetic $\op T$ and potential $\op V$ term, so that the equation of motion 
for $\ket{f_p(\omega)}$ is recasted as:
\be\lb{fluctuationStateEqMotion}
\left(\bbomega+\eps_p -\op T\right)\ket{f_p(\omega)} = \op V\ket{f_p(\omega)} + \op Q_0\left(\op V'(\omega) + \op \Phi(\omega)\right)\ket{\psi_p} \;.
\ee
This equation has the formal structure of an inhomogeneous Helmholtz equation and can be written in the coordinate representation as:
\be
\left(k^2_p(\omega)+\ii\eta + \nabla^2\right)f_p(\omega,\r) = s(\omega,\r) \;\;, {\rm with} \;\; k^2_p(\omega) = 2(\omega+\eps_p) \;.
\ee
The solution of this equation admits an integral representation in the form:
\be\lb{fluctuationStateIntEq1}
f_p(\omega,\r) = \int \dd\r' g_H(\r-\r')s(\r') \;,
\ee
where $g_H(\r)$ is the Green function of the Helmholtz operator in three dimensions. This integral representation allows us to assess the localization character 
of the fluctuation states, \emph{i.e.} the behavior of $f_p(\omega,\r)$ when $\r$ becomes large. Indeed, assuming that the support of the source term is compact
we have that, for $\r$ outside the domain in which the integrand of \eqref{fluctuationStateIntEq1} is not vanishing, the leading contribution of the fluctuation states
is given by:
\be\lb{fluctuationStateIntEq2}
f_p(\omega,\r) \sim \frac{e^{\ii \sqrt{k^2_p(\omega)+\ii\eta} |\r|}}{|\r|} \;.
\ee
We observe that the presence of the regularization parameter $\eta$ ensures a (slow) asymptotic convergence to zero, nonetheless the qualitative $\r$-dependence of the fluctuation 
states is determined by the sign of $k^2_p(\omega)$. In particular equation \eqref{fluctuationStateIntEq2} describes an exponential damping for $\omega < |\eps_p|$ and an oscillating 
power law behavior for $\omega \geq |\eps_p|$.
This argument holds for all the fluctuation states that contribute to the perturbation-induced expectation value \eqref{LinearResponseFunctDef3}, so we conclude that
the linear response functional is built in terms of localized fluctuation states for $\omega < |\eps_h|$. Instead, in the range $\eps_h|\leq\omega<|\eps_0|$ it contains both
contributions from local and non-local quantities and, finally is composed only by non-local fluctuation states for $\omega\geq|\eps_0|$.

\subsubsection{Criterion for the choice of the computational basis} %underlying

The capability of the computational basis to express the results of the linear response can be discussed on the localization argument previously presented. 
The first step needed to address this problem concerns with the evaluation of the reliability of the $\mu$-basis for $\omega=0$, \emph{i.e.} for a perturbing field that does 
not depend on time. In this case the fluctuation states, as well as the expectation value of a given observable, can be computed performing a ground state calculation in which the perturbing 
field $\op\Phi$ is added to the external potential of $\hnot$\footnote{We are interested in the linear response, so the magnitude of $\op\Phi$ has to be chosen in a way that respects this 
regime. To check this point we can compute the perturbation-induced modification of an observable an verify that its value scales linearly with the magnitude of the field}. 
The goodness of the $\mu$-basis can thus be verified as in the standard ground state calculation, for instance by checking that, given a certain level of completeness of the basis,
the energy of the occupied (perturbed) orbitals is stable with respect to further refinements.

Once that a basis with acceptable features has been chosen and verified for $\omega=0$ we can reasonably extend its reliability to $\omega>0$ by using the localization argument. 
Indeed, we have argued that fluctuation states retain a local behavior up to $\omega=|\eps_h|$ and so we can expect that, up to this threshold value, the $\mu$-basis represents an 
effective tool for a precise discretization of \eqref{LinearResponseFunctDef3}, thus providing a reliable estimate of the linear response.

@@

%\subsection{Excitation operator and the analytic structure of the linear response}
\subsection{Linear response \emph{via} the construction of the excitation operators}

@@

%\subsubsection{Definition of the basis of transitions}
\subsubsection{Explicit sampling of the empty-states subspace}

We introduce a set of basis vectors $\{\ket{\psi(\alpha)}\}$ in the subspace of the empty-states, \emph{i.e.} in the subspace orthogonal to the occupied eigenstates of $\hnot$. 
The elements of this set satisfy, by definition, the following orthonormality relation: 
\be\lb{psiAlphaDef1}%{orthonormalityDef1}
\brket{\psi(\alpha)}{\psi_p} = \brket{\psi_p}{\psi(\alpha)} = 0 \;, \qq
\brket{\psi(\alpha)}{\psi(\beta)} = \delta(\alpha-\beta) \;,
\ee
where we have parametrized the basis element through the multi-index $\alpha$, in the sense described in \ref{BasicDef}, since the associated vector can describe both a localized or a 
de-localized state. The $\op Q_0$ projector as well as the completeness relation can be easily written in function of this basis: 
\be\lb{psiAlphaQ0}
\op Q_0 = \sint\dd\alpha\ketbra{\psi(\alpha)}{\psi(\alpha)} \;, \;\; 
\identity = \sum_p\ketbra{\psi_p}{\psi_p} + \sint\dd\alpha\ketbra{\psi(\alpha)}{\psi(\alpha)} \;.
\ee
The dimension $n_\alpha$ of the $\alpha$-basis can be determined by expressing its elements $\{\ket{\psi(\alpha)}\}$ in the same $\mu$-basis used to perform the ground states calculation. 
%both with and without the explicit time-independent perturbing field. 
%Denoting with $n_{occ}$ the number of occupied orbitals  
Since the orthogonality condition with respect to the occupied orbitals imposes $n_{occ}$ constraints we conclude that $n_\alpha=n_\mu-n_{occ}$.

% We now introduce the notion of \emph{basis of transitions} and identify the operators than can be expanded in such a basis. 
% 
% We split the complete
% set of eigenstates of $\hnot$ in two sets. The first one is composed by the eigenvectors of $\hnot$ in the subspace of the occupied orbitals. The second 
% one is composed by the remaining eigenvectors, both in the bound and in the continuum region, and its elements are denoted as $\ket{\psi(\alpha)}$, where $\alpha$ 
% is a multi-index in the sense described in \ref{BasicDef}. The elements of these sets satisfy the following orthonormality relation: 
% \be\lb{orthonormalityDef1}
% \brket{\psi(\alpha)}{\psi_p} = \brket{\psi_p}{\psi(\alpha)} = 0 \;, \qq \brket{\psi_p}{\psi_q} = \delta_{pq} \;, \qq
% \brket{\psi(\alpha)}{\psi(\beta)} = \delta(\alpha-\beta) \;.
% \ee
% Obviously, the direct sum of these two sets coincides with the complete set of eigenstates of $\hnot$ and thus 
% constitute a complete basis.

\subsubsection{Linear response in the basis of transitions}

A generic operator can be decomposed in this basis and its formal expansion contains four distinct contribution of the form $\ketbra{\psi_p}{\psi_q}$, 
$\ketbra{\psi(\alpha)}{\psi(\beta)}$, $\ketbra{\psi(\alpha)}{\psi_p}$, $\ketbra{\psi_p}{\psi(\alpha)}$. However an inspection of equations \eqref{rhoPrimeOmegaDef1}
and \eqref{LiouvillianRhopomegaDef3} evidences that the relevant operators in the contest of the linear response are expressed as a commutator with $\dmnot$. 
For this class of operators the only non vanishing contribution comes from the \emph{transition operators}:
\be
\excite{p}{\alpha} \equiv \ketbra{\psi(\alpha)}{\psi_p}\;, \qq
\decay{\alpha}{p} \equiv \ketbra{\psi_p}{\psi(\alpha)} \;.
\ee
The orthonormality conditions \eqref{psiAlphaDef1} imply that:
\be\lb{transitionOpOrthonormalityDef1}
\trace{\excite{p}{\alpha}\excite{q}{\beta}} = \trace{\decay{\alpha}{p}\decay{\beta}{q}} = 0 \;, \qq
\trace{\excite{p}{\alpha}\decay{\beta}{q}} = \delta_{pq}\delta(\alpha-\beta) \;,
\ee
so that the expansion of a suitable operator $\op O$ in the basis of transitions reads:
\be\lb{OperatorTransBasisExpDef1}
\op O = \sum_p\sint \dd \alpha\left(
\excite{p}{\alpha}\trace{\decay{\alpha}{p}\op O} + \decay{\alpha}{p}\trace{\excite{p}{\alpha}\op O}
\right) \;.
\ee
In order to recast this expansion in a simpler expression it turns out to be useful to define the rank-2 row vectors of the transition basis: 
\be
\sbra{p\alpha}{\cdot} \equiv 
\mat{ \excite{p}{\alpha} \;, & \decay{\alpha}{p} }\;,
\ee
and we also introduce a vectorial notation to deal with the coefficients of the expansion of an operator in the transition basis, that is: 
\be\lb{OperatorTransitionCoeffDef1}
\sket{p\alpha}{\op O}  \equiv \mat{ \trace{\decay{\alpha}{p}\op O } \\ 
\trace{\excite{p}{\alpha}\op O}} \; .
\ee
Thanks to this formalism we have that equation \eqref{OperatorTransBasisExpDef1} can be expressed as:
\be\lb{OperatorTransBasisExpDef2}
\op O = \sum_p\sint \dd \alpha \sbra{p\alpha}{\cdot}\sket{p\alpha}{\op O} = 
\dbraket{\cdot}{\op O} \;,
\ee
where the last equality implicitly define the ulterior compact notation of the ``big-shaped'' vectors in which the summed indexes are omitted.  
In the same fashion it will be also useful for the subsequent analysis to introduce a ``conjugate'' notation, in which the basis operators are expressed
as column vector and the coefficients of the expanded operator as row ones, that is: 
\be\lb{TransitionBasisConjugateDef1}
\sket{p\alpha}{\cdot} \equiv 
\mat{\decay{\alpha}{p} \\  \excite{p}{\alpha}} \;, \qq
\sbra{p\alpha}{\op O}  \equiv 
\mat{ \trace{\excite{\alpha}{p}\op O } \;, & \trace{\decay{p}{\alpha}\op O} }
\;,
\ee
where the flip in the role of $\op e$ and $\op d$ in \eqref{TransitionBasisConjugateDef1} follows from the properties of excitation and decay operators
under hermitian-conjugate operation. The expansion of a suitable operator in the transition basis using the conjugate description reads: 
\be\lb{OperatorTransBasisConjugateExpDef1}
\op O = \sum_p\sint \dd \alpha \sbra{p\alpha}{\op O}\sket{p\alpha}{\cdot} = 
\dbraket{\op O}{\cdot} \;.
\ee

%\subsubsection{Spectral decomposition of the response density in the transitions basis}
